---
title: "Bike Share Analysis Udacity Project 2"
author: "Emmanuel Teikutey"
date: "9/29/2021"
output: pdf_document
---

# INTRODUCTION - BIKE SHARE DATA

Over the past decade, bicycle-sharing systems have been growing in number and popularity in cities across the world. Bicycle-sharing systems allow users to rent bicycles on a very short-term basis for a price. This allows people to borrow a bike from point A and return it at point B, though they can also return it to the same location if they'd like to just go for a ride. Regardless, each bike can serve several users per day.

Thanks to the rise in information technologies, it is easy for a user of the system to access a dock within the system to unlock or return bicycles. These technologies also provide a wealth of data that can be used to explore how these bike-sharing systems are used.

#PROJECT OVERVIEW

In this project, I will make use of R to explore and visualise data related to bike share systems for three major cities in the United States—Chicago, New York City, and Washington. I will write code to import the data and answer interesting questions about it by computing descriptive statistics and making visualizations!

## Major Questions to Answer

There are a number of different areas of interest to explore in the data set but for the purpose of this Project I will be focusing on the following questions:
1. The most common month of travel in the cities
2. The most common Start Station in the cities
3. Total travel time of users in the different cities.

# THE DATA SET

Randomly selected data for the first six months of 2017 are provided for all three cities.

Before I start answering the questions, let check the nature of the data set involved.


```{r cities}
ny = read.csv('new-york-city.csv')
wash = read.csv('washington.csv')
chi = read.csv('chicago.csv')
```
```{r}
head(ny)
```

```{r}
head(wash)
```
```{r}
head(chi)
```
```{r}
dim(ny)
```
```{r}
dim(wash)
```
```{r}
dim(chi)
```
All three of the data files contain 300000 rows and 10 columns for Chicago and NewYork whiles Washington has 8 columns. But all data files have same core six (6) columns:

Start Time (e.g., 2017-01-01 00:07:57)
End Time (e.g., 2017-01-01 00:20:53)
Trip Duration (in seconds - e.g., 776)
Start Station (e.g., Broadway & Barry Ave)
End Station (e.g., Sedgwick St & North Ave)
User Type (Subscriber or Customer)

The Chicago and New York City files also have the following two columns:

Gender
Birth Year

# DATA ANALYSIS/ ANSWER TO QUESTIONS

## QUESTION ONE: The most common month of travel in the cities

The date and time is grouped in the datetime. The montths are also numerically labelled and I have relabel them characteristically. I need to extract the month from the date time and to do this I need to use the lubrdate function which provides tools that make it easier to parse and manipulate dates.

```{r}
library(lubridate)

month <- function(city) {
  start_date <-
    sapply(strsplit(as.character(city$Start.Time), " "), "[", 1)
  # extract month from the datatime
  month <- substr(x = start_date, 6, 7)
}
# Call the month function to Chicago
month_chi <- month(chi)
# set a new column with the month data
chi['month'] <- month_chi
# list the month column to check when needed
#chi['month']

# Call the month function to New York when needed
month_ny <- month(ny)
ny['month'] <- month_ny
#ny['month']

# Call the month function to Washington when needed
month_wash <- month(wash)
wash['month'] <- month_wash
#wash['month']


# replacing '01' with 'january' and '02' with 'February' and so on ...
old <- c('01', '02', '03', '04', '05', '06')
new <- c('January', 'Febraury', 'March', 'April', 'May', 'June')
chi$month[chi$month %in% old] <-
  new[match(chi$month, old, nomatch = 0)]
ny$month[ny$month %in% old] <-
  new[match(ny$month, old, nomatch = 0)]
wash$month[wash$month %in% old] <-
  new[match(wash$month, old, nomatch = 0)]

# find the unique values of months
# check the list when needed
#chi['month']
uniqv_chi <- unique(chi$month)
#ny['month']
uniqv_ny <- unique(ny$month)
#wash['month']
uniqv_wash <- unique(wash$month)

# check the list when needed
#uniqv_chi
#uniqv_ny
#uniqv_wash

# find the mode of the month
common_month <- function(data_column,uniqv) {
  uniqv[which.max(tabulate(match(data_column, uniqv)))]   
}

# call the most common month function to Chicago
common_m_chi <- common_month(chi$month,uniqv_chi)
cat('The most common month for Chicago is:', common_m_chi,'\n')

```
```{r}
# call the most common month function to New York
common_m_ny <- common_month(ny$month,uniqv_ny)
cat('The most common month for New York is:', common_m_ny,'\n')
```
```{r}
# call the most common month function to Washington
common_m_wash <- common_month(wash$month,uniqv_wash)
cat('The most common month for Washington is:', common_m_wash)
```
 In all cases/ cities, it is evidently clear that June is the most common month.
 
 
## Visualisation: The most common of month in the cities

For visualization I chose to use ggplot2 function to create my graphs.

```{r, echo=FALSE}
chi$month <- factor(chi$month,levels = c('January', 'Febraury', 'March', 'April', 'May', 'June'))
ggplot(aes(chi$month),data=chi)+
  geom_bar(color='black',fill='#099DD9')+
  labs(x='Months',y='Counts for each month',title='Usage of Each Month in Chicago')
```

```{r}
ny$month <- factor(ny$month,levels = c('January', 'Febraury', 'March', 'April', 'May', 'June'))
ggplot(aes(ny$month),data=ny)+
  geom_bar(color='black',fill= 'red')+
  labs(x='Months',y='Counts for each month',title='Usage of Each Month in New York')
```
```{r}
wash$month <- factor(wash$month,levels = c('January', 'Febraury', 'March', 'April', 'May', 'June'))
ggplot(aes(wash$month),data=wash)+
  geom_bar(color='black',fill='green')+
  labs(x='Months',y='Counts for each month',title='Usage of Each Month in Washington')+
  scale_x_discrete(limits= c('January', 'Febraury', 'March', 'April', 'May', 'June'))
```
### Numeric Summary
```{r}
summary(chi$month)
```
Numeric summary for Chicago

```{r}
summary(ny$month)
```
Numeric summary for New York
```{r}
summary(wash$month)
```
Numeric summary for Washington
I all cases it is evidently clear the June was the most common Month

## QUESTION 2: What is the Most Common Start Station?

To answer this question, I need to assign unique values and find the mode for the various start stations in the cites.
```{r}
uniqv_ss_chi <- unique(chi$Start.Station)
uniqv_ss_ny <- unique(ny$Start.Station)
uniqv_ss_wash <- unique(wash$Start.Station)
```
```{r}
common_ss <- function(data_column,uniqv_ss) {
  uniqv_ss[which.max(tabulate(match(data_column, uniqv_ss)))]   
}
```

Call the most common month function to Chicago
```{r}
common_ss_chi <- common_ss(chi$Start.Station,uniqv_ss_chi)
print('The most common start station for Chicago is:') 
common_ss_chi
```
Call the most common month function to New York
```{r}
common_ss_ny <- common_ss(ny$Start.Station,uniqv_ss_ny)
print('The most common start station for New York is:') 
common_ss_ny
```
Call the most common month function to Washington
```{r}
common_ss_wash <- common_ss(wash$Start.Station,uniqv_ss_wash)
print('The most common start station for Washington is:') 
common_ss_wash
```
## VISUALISATION: Most Common Start Station
Plot for Chicago
```{r}
library(ggplot2)
library(scales)

ggplot(aes(chi$Start.Station),data=chi)+
  geom_bar(color='black',fill='#099DD9')+
  labs(x='Stations',y='Counts for each station',title='All Starting Stations in Chicago')
```
Plot for New York
```{r}
ggplot(aes(ny$Start.Station),data=ny)+
  geom_bar(color='black',fill='#099DD9')+
  labs(x='Stations',y='Counts for each station',title='Starting Stations in New York')
```
Plot for Washington
```{r}
ggplot(aes(wash$Start.Station),data=wash)+
  geom_bar(color='black',fill='#099DD9')+
  labs(x='Stations',y='Counts for each station',title='Starting Stations in Washington')
```

## QUESTION 3: What is the average travel time for users in different cities?

To get the average travel time for users, I run the mean function on the Trip duration column for the different cities.

```{r}
duration_chi = mean(chi$Trip.Duration)
duration_ny = mean(ny$Trip.Duration)
duration_wash = mean(wash$Trip.Duration)
```

Average time travel for Chicago
```{r}
cat('The average travel time for Chicago is:', duration_chi,'\n')
```
Average time travel for New York
```{r}
cat('The average travel time for New York is:', duration_ny,'\n')
```

Average time travel for New York
```{r}
cat('The average travel time for Washington is:', duration_wash,'\n')
```
## VISUALISATION: Average time travel for cities

Plot for Chicago
```{r}
ggplot(aes(x=Trip.Duration),data=chi)+
  geom_histogram(binwidth = 100)+
  ggtitle('The Bar Plot of Average Travel Time of Chicago')+
  scale_x_continuous(limits = c(0,5000))+
  labs(x='Usages',y='Travel Time')+
  geom_hline(aes(yintercept = mean(Trip.Duration)),col='red',size=1)
```

Plot for New York
```{r}
ggplot(aes(x=Trip.Duration),data=ny)+
  geom_histogram(binwidth = 100)+
  ggtitle('The Bar Plot of Average Travel Time of New York')+
  scale_x_continuous(limits = c(0,5000))+
  labs(x='Usages',y='Travel Time')+
  geom_hline(aes(yintercept = mean(Trip.Duration)),col='red',size=1)
```

Plot for Washington
```{r}
ggplot(aes(x=Trip.Duration),data=ny)+
  geom_histogram(binwidth = 100)+
  ggtitle('The Bar Plot of Average Travel Time of Washington')+
  scale_x_continuous(limits = c(0,5000))+
  labs(x='Usages',y='Travel Time')+
  geom_hline(aes(yintercept = mean(Trip.Duration)),col='red',size=1)
```

#OBESERVATIONS
- June is the most common months amongst the months under consideration
- The city with the highest travel time is New york
- The most common start station for Chicago is Street Dr & Grand Av, the one for New York is Perching Square North and Washington is Columbus Circle/ Union Station

# RECOMMENDATIONS
- More promotional activities should be carried out in other months other than June 
- Offer discounted pricing in areas with low patronage
